---
title: "Getting Started with wpgdata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with wpgdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE
)
```

## Introduction

`wpgdata` provides a simple R interface to query and download datasets
from the [City of Winnipeg Open Data Portal](https://data.winnipeg.ca)
using the OData V4 API. This vignette walks through the core functions
using the Assessment Parcels dataset (`d4mq-wa44`) as a working example.
```{r library}
library(wpgdata)
library(dplyr)
```

---

## Workflow overview

The typical `wpgdata` workflow follows four steps:
```
peg_info()      →   peg_metadata()   →   peg_query()   →   peg_all()
(explore)           (find fields)        (filter)          (download all)
```

---

## Step 1 — Explore a dataset with `peg_info()`

Before downloading data, use `peg_info()` to understand what a dataset
contains — its name, description, category, update frequency, and
row count:
```{r peg_info}
peg_info("d4mq-wa44")
```

This tells us the dataset was last updated, how many times it has been
downloaded, and what license it is published under — all useful before
committing to a large download.

---

## Step 2 — Find field names with `peg_metadata()`

OData queries require exact field names. Use `peg_metadata()` to look
up the available fields and their types before writing any query:
```{r peg_metadata}
meta <- peg_metadata("d4mq-wa44")
meta
```

The `field_name` column contains the names to use in `peg_query()`.
The `type` column tells you whether a field is text, number, or
another type — important for writing correct filter expressions.

For example, to find numeric fields only:
```{r meta_numeric}
meta |>
  dplyr::filter(type == "number")
```

---

## Step 3 — Fetch a quick preview with `peg_get()`

`peg_get()` fetches the first page of a dataset (up to the server's
default page size). It is useful for a quick look at the data
structure before writing more targeted queries:
```{r peg_get}
df <- suppressWarnings(peg_get("d4mq-wa44"))
df
```

Check column names and types:
```{r peg_get_glimpse}
dplyr::glimpse(df)
```

---

## Step 4 — Query with `peg_query()`

`peg_query()` supports five OData parameters:

| Argument | OData parameter | Purpose |
|---|---|---|
| `filter` | `$filter` | filter rows |
| `select` | `$select` | choose columns |
| `top` | `$top` | limit rows returned |
| `skip` | `$skip` | skip rows (offset) |
| `orderby` | `$orderby` | sort results |

### Filtering rows

Pass R expressions directly — `wpgdata` translates them to OData:
```{r filter_r}
peg_query("d4mq-wa44",
  filter = total_assessed_value > 1000000,
  top    = 10
)
```

Or use raw OData strings if you prefer:
```{r filter_odata}
peg_query("d4mq-wa44",
  filter = "total_assessed_value gt 1000000",
  top    = 10
)
```

Both approaches return identical results.

### Compound filters

Combine conditions with `&` (AND) and `|` (OR):
```{r filter_compound}
peg_query("d4mq-wa44",
  filter = total_assessed_value > 1000000 & building_type == "TWO STOREY",
  top    = 10
)
```

### Selecting columns

Use `select` to return only the columns you need:
```{r select}
peg_query("d4mq-wa44",
  select = c("roll_number", "full_address", "total_assessed_value",
             "building_type", "year_built"),
  top    = 10
)
```

### Sorting results

Use `orderby` to sort ascending or descending:
```{r orderby}
peg_query("d4mq-wa44",
  select  = c("roll_number", "full_address", "total_assessed_value"),
  orderby = "total_assessed_value desc",
  top     = 10
)
```

### Combining everything
```{r combined}
peg_query("d4mq-wa44",
  filter  = total_assessed_value > 1000000 & year_built > 2000,
  select  = c("roll_number", "full_address", "total_assessed_value",
              "building_type", "year_built"),
  orderby = "total_assessed_value desc",
  top     = 10
)
```

### Pagination with `skip`

Use `skip` and `top` together to paginate manually:
```{r skip}
# rows 1-5
page_1 <- peg_query("d4mq-wa44", top = 5, skip = 0)

# rows 6-10
page_2 <- peg_query("d4mq-wa44", top = 5, skip = 5)
```

---

## Step 5 — Download all rows with `peg_all()`

`peg_all()` automatically paginates through all pages and returns a
single tibble. It shows a progress bar with row counts and percentage:
```{r peg_all_cap}
# fetch up to 3 pages (safety cap)
peg_all("d4mq-wa44", max_pages = 3)
```

For large datasets, the default `max_pages = 10` prevents accidental
downloads. Set `max_pages = Inf` to fetch everything:
```{r peg_all_inf, eval = FALSE}
# fetch all rows - may take several minutes for large datasets
peg_all("d4mq-wa44", max_pages = Inf)
```

---

## OData filter reference

| R expression | OData equivalent | Meaning |
|---|---|---|
| `x == 1` | `x eq 1` | equal |
| `x != 1` | `x ne 1` | not equal |
| `x > 1` | `x gt 1` | greater than |
| `x >= 1` | `x ge 1` | greater than or equal |
| `x < 1` | `x lt 1` | less than |
| `x <= 1` | `x le 1` | less than or equal |
| `x == 1 & y == 2` | `(x eq 1 and y eq 2)` | AND |
| `x == 1 \| y == 2` | `(x eq 1 or y eq 2)` | OR |
| `!x` | `not x` | NOT |

---

## Finding dataset IDs

Every dataset on the [City of Winnipeg Open Data Portal](https://data.winnipeg.ca)
has a unique ID. To find it:

1. Browse to [data.winnipeg.ca](https://data.winnipeg.ca)
2. Open any dataset
3. Click **API** → **OData V4**
4. Copy the last segment of the URL:
```
https://data.winnipeg.ca/api/odata/v4/d4mq-wa44
                                      ^^^^^^^^^^
                                      dataset ID
```

---

## Further reading

- [City of Winnipeg Open Data Portal](https://data.winnipeg.ca)
- [Socrata OData Documentation](https://dev.socrata.com)
- [OData V4 Query Options](https://www.odata.org/documentation/)