---
title: "Getting Started with wpgdata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with wpgdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  cache     = TRUE
)
```

## Introduction

`wpgdata` provides a tidy R interface to the
[City of Winnipeg Open Data Portal](https://data.winnipeg.ca).
Discover available datasets, inspect their schemas, and download records
with automatic parallel pagination — all via the Socrata OData V4 and
Discovery APIs.

This vignette walks through the four core functions using the Assessment
Parcels dataset (`d4mq-wa44`) as the primary working example, with
additional examples using the 311 Service Requests dataset (`u7f6-5326`)
to demonstrate date filtering on large datasets.

```{r library}
library(wpgdata)
library(dplyr)
```

---

## Workflow overview

The typical `wpgdata` workflow follows four steps:

```
peg_catalogue()   →   peg_info()      →   peg_metadata()   →   peg_data()
(find datasets)       (dataset info)      (column schema)      (fetch rows)
```

| Function | Purpose |
|---|---|
| `peg_catalogue()` | Browse all available datasets and find IDs |
| `peg_info()` | Explore a specific dataset |
| `peg_metadata()` | Find field names before querying |
| `peg_data()` | Filter, select, sort, and download rows |

---

## Step 1 — Find datasets with `peg_catalogue()`

`peg_catalogue()` retrieves every published dataset from the portal.
Both catalogue pages and per-dataset metadata are fetched in parallel,
so the full catalogue arrives quickly regardless of size.

```{r peg_catalogue}
catalogue <- peg_catalogue()
catalogue
```

Count datasets by category to get an overview of what's available:

```{r peg_catalogue_count}
catalogue |>
  count(category, sort = TRUE)
```

Search by name to find a dataset and retrieve its ID:

```{r peg_catalogue_search}
catalogue |>
  filter(grepl("assessment", name, ignore.case = TRUE)) |>
  select(name, id, rows_updated_at)
```

Use the `id` value in any other `peg_*` function:

```{r peg_catalogue_id}
dataset_id <- catalogue |>
  filter(name == "Assessment Parcels") |>
  pull(id)

dataset_id
```

Use `limit` to cap results while exploring:

```{r peg_catalogue_limit}
peg_catalogue(limit = 10)
```

---

## Step 2 — Explore a dataset with `peg_info()`

`peg_info()` returns high-level metadata for a single dataset — name,
description, category, update frequency, row count, and license:

```{r peg_info}
peg_info("d4mq-wa44")
```

This is useful before committing to a large download: it tells you when
the data was last updated, how many rows to expect, and what license it
is published under.

---

## Step 3 — Find field names with `peg_metadata()`

OData queries require exact field names. Use `peg_metadata()` to look
up available fields and their types before writing any query:

```{r peg_metadata}
meta <- peg_metadata("d4mq-wa44")
meta
```

The `field_name` column contains the names to use in `peg_data()`.
The `type` column tells you whether a field is text, number, floating
timestamp, or another type — important for writing correct filter
expressions.

Find numeric fields only:

```{r meta_numeric}
meta |>
  filter(type == "number")
```

Find timestamp fields (relevant for date filtering):

```{r meta_timestamp}
meta |>
  filter(type == "calendar_date")
```

---

## Step 4 — Fetch data with `peg_data()`

`peg_data()` is the single function for fetching rows. It supports
server-side filtering, column selection, sorting, and offset pagination.
All pages are fetched in parallel automatically — no manual pagination
needed.

### Quick preview

Fetch a small sample to inspect structure before a larger query:

```{r peg_data_preview}
peg_data("d4mq-wa44", top = 5)
```

### Filtering rows

Pass R expressions directly — `wpgdata` translates them to OData syntax:

```{r filter_r}
peg_data("d4mq-wa44",
  filter = total_assessed_value > 1000000,
  top    = 10
)
```

Or use raw OData strings if you prefer:

```{r filter_odata}
peg_data("d4mq-wa44",
  filter = "total_assessed_value gt 1000000",
  top    = 10
)
```

Both approaches produce identical results.

### Compound filters

Combine conditions with `&` (AND) and `|` (OR):

```{r filter_compound}
peg_data("d4mq-wa44",
  filter = total_assessed_value > 1000000 & building_type == "TWO STOREY",
  top    = 10
)
```

### Selecting columns

Use `select` to return only the columns you need — reduces transfer size
significantly on wide datasets:

```{r select}
peg_data("d4mq-wa44",
  select = c("roll_number", "full_address", "total_assessed_value",
             "building_type", "year_built"),
  top    = 10
)
```

### Sorting results

Use `orderby` to sort ascending or descending:

```{r orderby}
peg_data("d4mq-wa44",
  select  = c("roll_number", "full_address", "total_assessed_value"),
  orderby = "total_assessed_value desc",
  top     = 10
)
```

### Combining filter, select, and orderby

```{r combined}
peg_data("d4mq-wa44",
  filter  = total_assessed_value > 1000000 & year_built > 2000,
  select  = c("roll_number", "full_address", "total_assessed_value",
              "building_type", "year_built"),
  orderby = "total_assessed_value desc",
  top     = 10
)
```

### Pagination with `skip`

Use `skip` and `top` together to retrieve a specific slice of rows:

```{r skip}
# rows 1–5
page_1 <- peg_data("d4mq-wa44", top = 5, skip = 0)

# rows 6–10
page_2 <- peg_data("d4mq-wa44", top = 5, skip = 5)
```

### Fetching all rows

Omit `top` to fetch every matching row across all pages:

```{r peg_data_all, eval = FALSE}
# fetches all rows — may take several minutes for large datasets
peg_data("d4mq-wa44")
```

---

## Date filtering

Datasets with timestamp columns (type `calendar_date` or
`floating_timestamp` in `peg_metadata()`) support date range filters
via raw OData strings. Socrata floating timestamps use ISO 8601 format
with milliseconds: `YYYY-MM-DDTHH:MM:SS.mmm`.

Build cutoff strings with a small helper to avoid repeating the format:

```{r ts_helper}
# helper — format a Date as a Socrata floating timestamp string
ts <- function(date) format(date, "%Y-%m-%dT00:00:00.000")
```

The examples below use the 311 Service Requests dataset (`u7f6-5326`),
which has 18+ million rows and is updated continuously — date filters
are essential to avoid downloading the entire dataset.

### Records closed in the last 4 days

```{r date_last4}
filter_str <- paste0("closed_date ge '", ts(Sys.Date() - 4), "'")

peg_data(
  "u7f6-5326",
  filter = filter_str,
  select = c("case_id", "channel_type", "subject", "reason",
             "type", "open_date", "closed_date", "case_status",
             "neighbourhood", "ward"),
  top    = 100L
)
```

### Records closed within a specific date range

```{r date_range}
filter_str <- paste0(
  "closed_date ge '", ts(as.Date("2026-03-01")), "'",
  " and ",
  "closed_date lt '", ts(as.Date("2026-03-07")), "'"
)

peg_data(
  "u7f6-5326",
  filter = filter_str,
  select = c("case_id", "subject", "open_date", "closed_date", "case_status"),
  top    = 100L
)
```

### Open cases from the last 7 days

```{r date_open}
filter_str <- paste0(
  "open_date ge '", ts(Sys.Date() - 7), "'",
  " and ",
  "case_status eq 'Open'"
)

peg_data(
  "u7f6-5326",
  filter  = filter_str,
  select  = c("case_id", "subject", "reason", "open_date",
              "case_status", "neighbourhood", "ward"),
  top     = 100L
)
```

### Cases closed yesterday, sorted by most recently closed

```{r date_yesterday}
filter_str <- paste0(
  "closed_date ge '", ts(Sys.Date() - 1), "'",
  " and ",
  "closed_date lt '", ts(Sys.Date()), "'"
)

peg_data(
  "u7f6-5326",
  filter  = filter_str,
  select  = c("case_id", "subject", "channel_type",
              "open_date", "closed_date", "neighbourhood"),
  orderby = "closed_date desc",
  top     = 100L
)
```

> **Note:** The `top` argument is included in all date filter examples
> to keep live API calls lightweight during development. Remove it to
> retrieve the full result set.

---

## OData filter reference

| R expression | OData equivalent | Meaning |
|---|---|---|
| `x == 1` | `x eq 1` | equal |
| `x != 1` | `x ne 1` | not equal |
| `x > 1` | `x gt 1` | greater than |
| `x >= 1` | `x ge 1` | greater than or equal |
| `x < 1` | `x lt 1` | less than |
| `x <= 1` | `x le 1` | less than or equal |
| `x == 1 & y == 2` | `(x eq 1 and y eq 2)` | AND |
| `x == 1 \| y == 2` | `(x eq 1 or y eq 2)` | OR |
| `!x` | `not x` | NOT |

> **Tip:** R expression syntax works for numeric and string comparisons.
> For date comparisons, use raw OData strings as shown in the date
> filtering section above.

---

## Finding dataset IDs

The easiest way is to search directly in R:

```{r find_id, eval = FALSE}
peg_catalogue() |>
  filter(grepl("your search term", name, ignore.case = TRUE)) |>
  select(name, id, category)
```

Alternatively, browse the
[City of Winnipeg Open Data Portal](https://data.winnipeg.ca), open any
dataset, click **API → OData V4**, and copy the last segment of the URL:

```
https://data.winnipeg.ca/api/odata/v4/d4mq-wa44
                                      ^^^^^^^^^^
                                      dataset ID
```

---

## Further reading

- [City of Winnipeg Open Data Portal](https://data.winnipeg.ca)
- [Socrata Developer Portal](https://dev.socrata.com)
- [OData V4 Query Options](https://www.odata.org/documentation/)
